---
title: "sc_mixology dataset quality control using DropletQC"
author: "Valerio Reffo"
date: 30-06-2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup-libraries

library(DropletQC)
library(purrr)
library(ggplot2)
library(patchwork)
library(dplyr)
library(scuttle)
library(scran)
library(scater)
```

## Data Loading and Nuclear Fraction Calculation

We import the main Mixology `SingleCellExperiment` dataset and calculate the per-cell nuclear fraction metric dynamically using the ratio of exonic mapping sequencing reads.

```{r}
#| label: load-sce

# Load the main mixology SCE object
sc_mix <- readRDS("/projects/shared/intronic_bam/datasets/sc_mixology/sc_mix.RDS")
sc_mix
```

### Exonic Fraction Processing and Metric Annotation

We inspect the global exonic mapping proportions and compute the true total sequenced reads per cell to accurately derive the nuclear fraction metric.

```{r}
#| label: calculate-nuclear-fraction
#| eval: false
#| include: false

# Inspect baseline exonic metrics
hist(prop.table(sc_mix$mapped_to_exon))
summary(sc_mix$mapped_to_exon)
head(colData(sc_mix))

# Calculate the total raw sequenced reads for each individual cell
total_sequenced_reads <- colData(sc_mix)$unaligned + 
  colData(sc_mix)$aligned_unmapped + 
  colData(sc_mix)$mapped_to_exon + 
  colData(sc_mix)$mapped_to_intron + 
  colData(sc_mix)$ambiguous_mapping + 
  colData(sc_mix)$mapped_to_MT + 
  colData(sc_mix)$mapped_to_ERCC

# Calculate the precise nuclear fraction and store it in colData
colData(sc_mix)$nuclear_fraction <- (colData(sc_mix)$mapped_to_exon / total_sequenced_reads)
```

---

## Exploratory Analysis

We visualize the distribution of our newly calculated nuclear fraction across the distinct cell lines and explore its scaling relationship with total UMI counts.

```{r}
#| label: plot-distributions
#| fig-width: 10
#| fig-height: 6

# Nuclear fraction distribution profile faceted by cell line
p <- ggplot(as.data.frame(colData(sc_mix)), aes(x = nuclear_fraction))

p + geom_density(fill = "lightblue", alpha = 0.5) +
  facet_wrap(~cell_line) +
  theme_minimal() +
  labs(title = "Distribution of Nuclear Fraction for Each Cell Type - Mixology Dataset",
       x = "Nuclear Fraction") +
  xlim(c(0, 1))

# UMI counts vs Nuclear Fraction distribution profile
p + aes(y = log10(sum), colour = cell_line) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_minimal() +
  labs(title = "UMI Counts vs Nuclear Fraction by Cell Line",
       x = "Nuclear Fraction",
       y = "log10(UMI Sum)")
```

---

## Normalization and Dimensionality Reduction

We perform log-transformation normalization on the raw counts, model features to extract the Top 10% Highly Variable Genes (HVGs), and generate low-dimensional embeddings.

```{r}
#| label: dimensionality-reduction
#| eval: false
#| include: false

sc_mix <- logNormCounts(sc_mix)
dec <- modelGeneVar(sc_mix)
hvgs <- getTopHVGs(dec, prop = 0.1)

sc_mix <- runPCA(sc_mix, subset_row = hvgs, ncomponents = 50)
sc_mix <- runUMAP(sc_mix, dimred = "PCA")
sc_mix
```

### Visualizing Embeddings (PCA vs UMAP)

We display the low-dimensional projection spaces side-by-side, assessing separation performance across cell lines and variations driven by the nuclear fraction values.

```{r}
#| label: plot-dimred
#| fig-width: 12
#| fig-height: 10

umapp <- plotUMAP(sc_mix, colour_by = "cell_line") + theme(legend.position = "bottom")
pcap  <- plotPCA(sc_mix, colour_by = "cell_line") + theme(legend.position = "none")

umapb <- plotUMAP(sc_mix, colour_by = "nuclear_fraction") + theme(legend.position = "bottom")
pcab  <- plotPCA(sc_mix, colour_by = "nuclear_fraction") + theme(legend.position = "none")

# Composite multi-panel layout
(pcap | pcab) / (umapp | umapb)
```

---

## Detecting Empty Droplets and Damaged Cells with `DropletQC`

We compile the nuclear fraction values and total UMI counts into a structural dataframe, filter out empty observations, and compute the cell status profiles.

```{r}
#| label: empty-droplets
#| fig-width: 8
#| fig-height: 6

# Prepare the data structure for the empty droplet detection algorithm
nf_umi <- data.frame(
  nf = sc_mix$nuclear_fraction,
  umi = sc_mix$sum
)


# Evaluate call and output diagnostic empty droplet cut-off visualization
cb_mix <- identify_empty_drops(nf_umi = nf_umi, include_plot = TRUE)
```

### Damaged Cell Identification

We isolate cells suffering from experimental stress or cellular damage by aligning properties against our benchmarked cell lines.

```{r}
#| label: damaged-cells
#| fig-width: 10
#| fig-height: 12

# Reorder metadata columns with dplyr relocate
cb_mix$cell_line <- sc_mix$cell_line

# Identify cell-damage status using DropletQC
cb_mix_dc <- identify_damaged_cells(cb_mix, verbose = FALSE, output_plots = TRUE)

# Show classification output cross-tabulated across the populations
table(cb_mix_dc[[1]]$cell_line, cb_mix_dc[[1]]$cell_status)

# Generate multi-panel structural QC diagnostics plots
wrap_plots(cb_mix_dc[[2]], nrow = 5)
```

### Saving the Updated SCE Object

```{r}
#| label: save-object
#| eval: false
#| include: false

saveRDS(sc_mix, "/projects/shared/intronic_bam/datasets/sc_mixology/sc_mix.RDS")
```