---
title: "Cord Blood Quality Control with DropletQC"
author: "Valerio Reffo"
date: 30-06-2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(DropletQC)
library(purrr)
library(ggplot2)
library(patchwork)
library(dplyr)
library(scuttle)
library(scran)
library(scater)
library(anndataR)
library(SingleCellExperiment)
```

## Data Loading and Initial Per-Cell QC

We load the calculated nuclear fraction data along with the main Cord Blood `SingleCellExperiment` object. We also calculate standard mitochondrial QC metrics.

```{r}
#| label: load-data
#| eval: false
#| include: false

# Load nuclear fraction and main SCE object
nf <- readRDS("~/scFound_bench/results/nf_cordblood.rds")
```


```{r}
#| label: load-sce

sce_cb <- readRDS("/projects/shared/intronic_bam/datasets/cord_blood/cord_blood_sce.RDS")
sce_cb
```

Let's read the anndata with the embeddings matrix
```{r}
#| label: anndata-read
#| eval: false

# Path automatically adapted for Cord Blood
cb_ann <- read_h5ad("/projects/shared/intronic_bam/datasets/anndata/cb_scGPT_embeddings.h5ad")
cb_ann
```

Add it into the reducedDim slot of the sce
```{r}
#| label: embeddings-add
#| eval: false

scgpt_matrix <- as.matrix(cb_ann$obsm['X_scGPT']$X_scGPT)
scgpt_matrix_aligned <- scgpt_matrix[colnames(sce_cb), ]
reducedDim(sce_cb, "scGPT") <- scgpt_matrix_aligned
reducedDimNames(sce_cb)
```

### Mitochondrial Metrics Calculation

We identify mitochondrial genes based on the `MT-` prefix to compute per-cell quality control metrics.

```{r}
#| label: mito-qc
#| eval: false

# Identify mitochondrial genes based on rownames
is_mito <- grepl("^MT-", rownames(sce_cb), ignore.case = TRUE)
summary(is_mito) # Verification of detected mitochondrial genes

# Automatically calculate and append QC metrics to colData
sce_cb <- addPerCellQC(sce_cb, subsets = list(Mito = is_mito))
```

---

## Aligning and Integrating Nuclear Fraction

Before adding the nuclear fraction to the SCE object, we examine its global distribution and accurately filter it to match the exact barcodes present in our filtered dataset.

```{r}
#| label: integrate-nuclear-fraction
#| eval: false
#| fig-height: 5
#| fig-width: 12
#| include: false

# Check original dimensions and distribution
dim(nf)
hist(nf$nuclear_fraction, main = "Original Nuclear Fraction Distribution", xlab = "Nuclear Fraction")

# Extract rows corresponding only to valid cell barcodes
nf_filtered <- nf[colnames(sce_cb), , drop = FALSE]
hist(nf_filtered$nuclear_fraction, main = "Filtered Nuclear Fraction Distribution", xlab = "Nuclear Fraction")

# Append the filtered nuclear fraction to colData
sce_cb$nuclear_fraction_QC <- nf_filtered$nuclear_fraction
```


```{r}
#| label: cell-table

# Review cell type composition
table(sce_cb$celltype)
```

---

## Exploratory Analysis and Distributions

We visualize the distribution of the nuclear fraction across different cell types and evaluate its relationship with total UMI counts.

```{r}
#| label: plot-distributions
#| fig-width: 10
#| fig-height: 6

# Distribution of nuclear fraction per cell type
p <- ggplot(as.data.frame(colData(sce_cb)), aes(x = nuclear_fraction))

p + geom_density(fill = "lightgreen", alpha = 0.5) +
  facet_wrap(~celltype) +
  theme_minimal() +
  labs(title = "Distribution of Nuclear Fraction for Each Cell Type - Cord Blood Dataset",
       x = "Nuclear Fraction (QC)")

# Distribution of UMI counts mapped over the cell types
p + aes(y = log10(sum), colour = celltype) +
  geom_point(alpha = 0.7, size = 0.8) +
  theme_minimal() +
  labs(title = "UMI Counts vs Nuclear Fraction by Cell Type",
       x = "Nuclear Fraction (QC)",
       y = "log10(UMI Sum)")

```

---

## Normalization and Dimensionality Reduction

We perform log-normalization, model the gene variance to select the Top 10% highly variable genes (HVGs), and generate PCA and UMAP embeddings.

```{r}
#| label: dimensionality-reduction
#| eval: false

sce_cb <- logNormCounts(sce_cb)
dec <- modelGeneVar(sce_cb)
hvgs <- getTopHVGs(dec, prop = 0.1)

sce_cb <- runPCA(sce_cb, subset_row = hvgs, ncomponents = 50)
sce_cb <- runUMAP(sce_cb, dimred = "PCA")
sce_cb
```

Let's calculate the UMAP on the embeddings too
```{r}
#| label: embed-umap
#| eval: false

sce_cb <- runUMAP(sce_cb, dimred = "scGPT", name = "UMAP_embed")
```

### Embedding Visualization (PCA vs UMAP)

We display the low-dimensional spaces colored by cell type and by our updated nuclear fraction metric.

```{r}
#| label: plot-dimred
#| fig-width: 12
#| fig-height: 10

# Cell identity
umapp <- plotUMAP(sce_cb, colour_by = "celltype", dimred = "UMAP") + 
  theme(legend.position = "bottom") + 
  labs(title = "UMAP (PCA) colored by cell identity")
pcap  <- plotPCA(sce_cb, colour_by = "celltype") + 
  theme(legend.position = "none") +
  labs(title = "PCA colored by cell identity")

# Nuclear Fraction
umapb <- plotUMAP(sce_cb, colour_by = "nuclear_fraction", dimred = "UMAP") + 
  theme(legend.position = "bottom") +
  labs(title = "UMAP (PCA) colored by nuclear fraction")
pcab  <- plotPCA(sce_cb, colour_by = "nuclear_fraction") + 
  theme(legend.position = "none") +
  labs(title = "PCA colored by nuclear fraction")

# Umap embeddings
umapes <- plotUMAP(sce_cb, colour_by = "celltype", dimred = "UMAP_embed") + 
  theme(legend.position = "bottom") +
  labs(title = "UMAP (Embeddings) colored by cell identity")
umape <- plotUMAP(sce_cb, colour_by = "nuclear_fraction", dimred = "UMAP_embed") + 
  theme(legend.position = "bottom") +
  labs(title = "UMAP (Embeddings) colored by nuclear fraction")

# Composite layout
(pcap | pcab) / (umapp | umapb)

(umapes | umape)
```

---

## Investigating MALAT1 and Nuclear Fraction Relationship

MALAT1 is absent in cordblood

```{r}
#| label: malat1-plots
#| eval: false
#| fig-height: 5
#| fig-width: 12


# Check feature expression on UMAP and PCA
umapm <- plotUMAP(sce_cb, colour_by = "MALAT1")
pcam  <- plotPCA(sce_cb, colour_by = "MALAT1")

# Side-by-side comparison with nuclear fraction
pcab | pcam
umapb | umapm
```

### MALAT1 vs Nuclear Fraction Scatter Plot

```{r}
#| label: scatter-malat1
#| eval: false
#| fig-height: 6
#| fig-width: 12

plotExpression(sce_cb, 
               features = "MALAT1", 
               x = "nuclear_fraction", 
               other_fields = "celltype",
               exprs_values = "logcounts") +
  facet_wrap(~celltype) + 
  aes(color = celltype) + 
  theme_minimal() +
  labs(title = "MALAT1 vs Nuclear Fraction per Cell Type",
       x = "Nuclear Fraction (QC)",
       y = "MALAT1 Expression (logcounts)")
```

---

## Empty Droplet and Damaged Cell Identification

Using `DropletQC`, we extract the nuclear fraction and total UMI counts to identify and flag potential empty droplets.

```{r}
#| label: empty-droplets
#| fig-width: 8
#| fig-height: 6

# Prepare the input data frame
nf_umi <- data.frame(
  nf = sce_cb$nuclear_fraction,
  umi = sce_cb$sum,
  celltype = sce_cb$celltype
)

rownames(nf_umi) <- colnames(sce_cb)

# Handle NA values
nf_umi <- nf_umi[!is.na(nf_umi$nf),]

# Re-run to generate diagnostic plots and visualize the threshold cut-offs
cb_ed <- identify_empty_drops(nf_umi = nf_umi, include_plot = TRUE)
```

```{r}
#| label: cell-status-prop
# Summary table of cell status
table(cb_ed$cell_status)
```

## Investigation of possible association between empty droplets and cell identity

Add the cell status column in the coldata of sce_cb. Note that MALAT1 is absent in this dataset, so we only evaluate the nuclear fraction distribution.

```{r}
#| label: cell_status-add
#| eval: false
sce_cb$cell_status <- cb_ed[colnames(sce_cb), "cell_status"]
```

```{r}
table(sce_cb$celltype, sce_cb$cell_status)
```

```{r}
#| label: cell-status-boxplots
#| fig-width: 10
#| fig-height: 6

p1 <- sce_cb[, !is.na(sce_cb$cell_status)] |> 
  ggcells(aes(x = cell_status, y = nuclear_fraction, fill = cell_status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  
  facet_wrap(~ celltype) +                           
  theme_minimal() + 
  labs(
    title = "Nuclear fraction distribution for cell status",
    subtitle = "By each cell identity",
    x = "Cell Status",
    y = "Nuclear Fraction",
    fill = "Status"
  )

p1
```

---

## Damaged Cell Identification

Using the cell type annotations, we call `identify_damaged_cells` to pinpoint cell structural damage or lysis, inspect the targeted cell populations, and save the finalized dataset.

```{r}
#| label: damaged-cells
#| fig-width: 10
#| fig-height: 12

# Reorder columns with relocate using celltype
cb_ed <- cb_ed |> 
  relocate(celltype, .after = cell_status)

# Detect damaged cell populations
cb_ed_dc <- identify_damaged_cells(cb_ed, verbose = FALSE, output_plots = TRUE)

# Final cell status summary
table(cb_ed_dc[[1]]$cell_status)

# Diagnostic plots for each cell identity group
wrap_plots(cb_ed_dc[[2]], nrow = 6)
```

Add again in the single cell experiment

```{r}
#| label: cell_status-add2
#| eval: false
sce_cb$cell_status <- cb_ed_dc[[1]][colnames(sce_cb), "cell_status"]
```

```{r}
table(sce_cb$celltype, sce_cb$cell_status)
```

```{r}
#| label: cell-status-boxplots2
#| fig-width: 10
#| fig-height: 6

p1 <- sce_cb[, !is.na(sce_cb$cell_status)] |> 
  ggcells(aes(x = cell_status, y = nuclear_fraction, fill = cell_status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  
  facet_wrap(~ celltype) +                           
  theme_minimal() + 
  labs(
    title = "Nuclear fraction distribution for cell status",
    subtitle = "By each cell identity",
    x = "Cell Status",
    y = "Nuclear Fraction",
    fill = "Status"
  )

p1
```

### Exporting the Finalized Object

```{r}
#| label: save-dataset
#| eval: false
#| include: false

saveRDS(sce_cb, "/projects/shared/intronic_bam/datasets/cord_blood/cord_blood_sce.RDS")
```