---
title: "BE1 dataset quality control using DropletQC"
author: "Valerio Reffo"
date: 30-06-2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup-libraries

library(DropletQC)
library(purrr)
library(ggplot2)
library(patchwork)
library(dplyr)
library(scuttle)
library(scran)
library(scater)
```

## Data Loading and Nuclear Fraction Integration

We define the sample IDs and import the RDS files containing the nuclear fraction calculated for each individual sample, then map them back to the main `SingleCellExperiment` object.

```{r}
#| label: load-nf
#| eval: false
#| include: false

samples <- c("A549", "CCL-185-IG", "PC9", "CRL5868", "HTB178", "DV90", "HCC78", "PBMCs")

# 1. Load the list of nuclear fractions
nf_list <- imap(samples, \(sample, name) {
  s_path <- paste0("~/scFound_bench/results/nf_BE1_", sample, ".rds")
  readRDS(s_path)
})
names(nf_list) <- samples
```


```{r}
#| label: load-sce
# 2. Load the main SCE object
sce_be1 <- readRDS("/projects/shared/intronic_bam/datasets/BE1/BE1_sce.RDS")
sce_be1
```

### Mapping Nuclear Fraction to `colData`

We merge the individual nuclear fraction data frames into a single object, ensuring that the constructed cell keys perfectly match the `colnames` of our SCE object.

```{r}
#| label: merge-nuclear-fraction
#| eval: false
#| include: false

# Combine the list into a single data frame
nf_combined <- do.call(rbind, lapply(names(nf_list), function(s) {
  df <- as.data.frame(nf_list[[s]])
  # Create a cell key identical to sce_be1 colnames (e.g., "AAACCCAAGCCTATCA-1_A549")
  df$cell_key <- paste0(rownames(df), "_", s)
  return(df)
}))

# Safe mapping using match()
colData(sce_be1)$nuclear_fraction <- nf_combined$nuclear_fraction[match(colnames(sce_be1), nf_combined$cell_key)]

# Verify the integration
head(colData(sce_be1)[, c("Sample", "Barcode", "nuclear_fraction")])
table(sce_be1$Sample)
```

---

## Exploratory Analysis

We visualize the distribution of the nuclear fraction across different samples and explore its relationship with total UMI counts to identify potential ambient background or empty droplets.

```{r}
#| label: plot-distributions
#| fig-width: 10
#| fig-height: 6

# Nuclear fraction distribution by sample identity
p <- ggplot(as.data.frame(colData(sce_be1)), aes(x = nuclear_fraction))

p + geom_density(fill = "lightblue", alpha = 0.5) +
  facet_wrap(~Sample) +
  theme_minimal() +
  labs(title = "Distribution of Nuclear Fraction - BE1 Dataset by Cell Identity",
       x = "Nuclear Fraction")

# UMI counts vs Nuclear Fraction distribution
p + aes(y = log10(sum), colour = Sample) +
  geom_point(alpha = 0.4, size = 0.6) +
  theme_minimal() +
  labs(title = "UMI Counts vs Nuclear Fraction", 
       x = "Nuclear Fraction",
       y = "log10(UMI Sum)")
```

---

## Normalization and Dimensionality Reduction

We perform log-transformation on the counts, select the Top 10% Highly Variable Genes (HVGs), and compute PCA and UMAP coordinates.

```{r}
#| label: dimensionality-reduction

sce_be1 <- logNormCounts(sce_be1)
dec <- modelGeneVar(sce_be1)
hvgs <- getTopHVGs(dec, prop = 0.1)

sce_be1 <- runPCA(sce_be1, subset_row = hvgs, ncomponents = 50)
sce_be1 <- runUMAP(sce_be1, dimred = "PCA")
sce_be1
```

### Visualizing Embeddings (PCA vs UMAP)

We compare the embedding spaces side-by-side, coloring cells by Sample and Nuclear Fraction using `patchwork`.

```{r}
#| label: plot-dimred
#| fig-width: 12
#| fig-height: 10

umapp <- plotUMAP(sce_be1, colour_by = "Sample") + theme(legend.position = "bottom")
pcap  <- plotPCA(sce_be1, colour_by = "Sample") + theme(legend.position = "none")

umapb <- plotUMAP(sce_be1, colour_by = "nuclear_fraction") + theme(legend.position = "bottom")
pcab  <- plotPCA(sce_be1, colour_by = "nuclear_fraction") + theme(legend.position = "none")

# Composite layout
(pcap | pcab) / (umapp | umapb)
```

---

## Focus on MALAT1 and Nuclear Fraction Relationship

We evaluate the behavior of the nuclear-retained long non-coding RNA **MALAT1** to biologically validate our nuclear fraction metric.

```{r}
#| label: malat1-analysis
#| fig-width: 12
#| fig-height: 5

umapm <- plotUMAP(sce_be1, colour_by = "MALAT1")
pcam  <- plotPCA(sce_be1, colour_by = "MALAT1")

# Direct comparison in reduced dimensions
umapb | umapm
```

### MALAT1 vs Nuclear Fraction Scatter Plot Faceted by Sample

```{r}
#| label: scatter-malat1-nf
#| fig-width: 12
#| fig-height: 6

plotExpression(sce_be1, 
               features = "MALAT1", 
               x = "nuclear_fraction", 
               other_fields = "Sample",
               exprs_values = "logcounts") +
  facet_wrap(~Sample) + 
  aes(color = Sample) + 
  theme_minimal() +
  labs(title = "MALAT1 Expression vs Nuclear Fraction per Sample",
       x = "Nuclear Fraction",
       y = "MALAT1 Expression (logcounts)")
```

---

### Detecting Empty Droplets and Damaged cells with `DropletQC`

We prepare the data frame for the `DropletQC` algorithm by handling missing values (`NA`) and evaluating the empty droplet cutoff.

```{r}
#| label: empty-droplets
#| fig-width: 8
#| fig-height: 6

nf_umi <- data.frame(
  nf = sce_be1$nuclear_fraction,
  umi = sce_be1$sum,
  Sample = sce_be1$Sample
)

# Handle NA values
nf_umi <- nf_umi[!is.na(nf_umi$nf),]

# Identify empty droplets and output the diagnostic plot
cb_be1 <- identify_empty_drops(nf_umi = nf_umi, include_plot = TRUE)
```


```{r}
# Summary table of cell status
table(cb_be1$cell_status)
```

### Damaged Cell Identification

Finally, we identify damaged cells by leveraging cell type annotations and save the finalized object.

```{r}
#| label: damaged-cells
#| fig-width: 10
#| fig-height: 12

# Reorder columns
cb_be1 <- cb_be1 |>  
  relocate(Sample, .after = cell_status)

# Identify damaged cells using DropletQC
cb_be1_dc <- identify_damaged_cells(cb_be1, verbose = FALSE, output_plots = TRUE)

# Final cell status summary
table(cb_be1_dc[[1]]$cell_status)

# Diagnostic plots for each cell identity group
wrap_plots(cb_be1_dc[[2]], nrow = 6)
```

### Saving the Updated SCE Object

```{r}
#| label: save-object
#| eval: false
#| include: false

saveRDS(sce_be1, file = "/projects/shared/intronic_bam/datasets/BE1/BE1_sce.RDS")
```

