---
title: "BE1 dataset quality control using DropletQC"
author: "Valerio Reffo"
date: 30-06-2026
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(DropletQC)
library(purrr)
library(ggplot2)
library(patchwork)
library(dplyr)
library(scuttle)
library(scran)
library(scater)
library(anndataR)
library(SingleCellExperiment)
```

## Data Loading and Nuclear Fraction Integration

We define the sample IDs and import the RDS files containing the nuclear fraction calculated for each individual sample, then map them back to the main `SingleCellExperiment` object.

```{r}
#| label: load-nf
#| eval: false


samples <- c("A549", "CCL-185-IG", "PC9", "CRL5868", "HTB178", "DV90", "HCC78", "PBMCs")

# 1. Load the list of nuclear fractions
nf_list <- imap(samples, \(sample, name) {
  s_path <- paste0("~/scFound_bench/results/nf_BE1_", sample, ".rds")
  readRDS(s_path)
})
names(nf_list) <- samples
```

```{r}
#| label: load-sce

sce_be1 <- readRDS("/projects/shared/intronic_bam/datasets/BE1/BE1_sce.RDS")
sce_be1
```

#### scGPT embeddings

Let's read the anndata with the embeddings matrix

```{r}
#| label: anndata-read
#| eval: false

be_ann <- read_h5ad("/projects/shared/intronic_bam/datasets/anndata/be1_scGPT_embeddings.h5ad")
```

Add it into the reducedDim slot of the sce

```{r}
#| label: embeddings-add
#| eval: false

scgpt_matrix <- as.matrix(be_ann$obsm['X_scGPT']$X_scGPT)
scgpt_matrix_aligned <- scgpt_matrix[colnames(sce_be1), ]
reducedDim(sce_be1, "scGPT") <- scgpt_matrix_aligned
```

#### GeneFormer embeddings

```{r}
#| label: csv-emb-load
#| eval: false

emb_gf <-readr::read_csv("/projects/shared/intronic_bam/datasets/embeddings/geneformer/be1_geneformer.csv")
```

Add it into the reduced dims of the sce

```{r}
#| label: merge-gf-embs
#| eval: false

reducedDim(sce_be1, "GeneFormer") <- emb_gf[match(as.character(colnames(sce_be1)), as.character(emb_gf$Barcode)), ] |>
  select(where(is.numeric), -any_of(c("Unnamed: 0", "...1"))) |>
  as.matrix() |>
  `rownames<-`(as.character(colnames(sce_be1)))

```

### Mapping Nuclear Fraction to `colData`

We merge the individual nuclear fraction data frames into a single object, ensuring that the constructed cell keys perfectly match the `colnames` of our SCE object.

```{r}
#| label: merge-nuclear-fraction
#| eval: false

# Combine the list into a single data frame
nf_combined <- do.call(rbind, lapply(names(nf_list), function(s) {
  df <- as.data.frame(nf_list[[s]])
  # Create a cell key identical to sce_be1 colnames (e.g., "AAACCCAAGCCTATCA-1_A549")
  df$cell_key <- paste0(rownames(df), "_", s)
  return(df)
}))

# Safe mapping using match()
colData(sce_be1)$nuclear_fraction <- nf_combined$nuclear_fraction[match(colnames(sce_be1), nf_combined$cell_key)]

# Verify the integration
head(colData(sce_be1)[, c("Sample", "Barcode", "nuclear_fraction")])
table(sce_be1$Sample)
```

------------------------------------------------------------------------

## Exploratory Analysis

We visualize the distribution of the nuclear fraction across different samples and explore its relationship with total UMI counts to identify potential ambient background or empty droplets.

```{r}
#| label: plot-distributions
#| layout-ncol: 2
#| fig-cap:
#|   - "Distribution of Nuclear Fraction - BE1 Dataset by Cell Identity"
#|   - "UMI Counts vs Nuclear Fraction"

# Primo grafico: Distribuzione di densità
p1 <- ggplot(as.data.frame(colData(sce_be1)), aes(x = nuclear_fraction)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  facet_wrap(~Sample) +
  theme_minimal() +
  labs(
    x = "Nuclear Fraction"
  )

# Secondo grafico: Scatter plot UMI vs Nuclear Fraction
p2 <- ggplot(as.data.frame(colData(sce_be1)), aes(x = nuclear_fraction, y = log10(sum), colour = Sample)) +
  geom_point(alpha = 0.4, size = 0.6) +
  theme_minimal() +
  labs(
    x = "Nuclear Fraction",
    y = "log10(UMI Sum)"
  )

p1
p2
```

------------------------------------------------------------------------

## Normalization and Dimensionality Reduction

We perform log-transformation on the counts, select the Top 10% Highly Variable Genes (HVGs), and compute PCA and UMAP coordinates.

```{r}
#| label: dimensionality-reduction
#| eval: false

sce_be1 <- logNormCounts(sce_be1)
dec <- modelGeneVar(sce_be1)
hvgs <- getTopHVGs(dec, prop = 0.1)

sce_be1 <- runPCA(sce_be1, subset_row = hvgs, ncomponents = 50)
sce_be1 <- runUMAP(sce_be1, dimred = "PCA")
sce_be1
```

Let's calculate the UMAP on the embeddings of scGPT and GeneFormer

```{r}
#| label: umap-scgpt
#| eval: false

sce_be1 <- runUMAP(sce_be1, dimred = "scGPT", name = "UMAP_scGPT")
```

```{r}
#| label: umap-gf
#| eval: false

sce_be1 <- runUMAP(sce_be1, dimred = "GeneFormer", name = "UMAP_GeneFormer")
```

```{r}
# Rename 'Geneformer' to 'Geneformer_V2'
reducedDimNames(sce_be1)[reducedDimNames(sce_be1) == "UMAP_embed"] <- "UMAP_scGPT"
```

### Visualizing Embeddings (PCA vs UMAP)

We compare the embedding spaces side-by-side, coloring cells by Sample and Nuclear Fraction using `patchwork`.

```{r}
#| label: plot-dimred
#| layout-ncol: 2
#| fig-cap:
#|   - "Distribution of Nuclear Fraction - BE1 Dataset by Cell Identity"
#|   - "UMI Counts vs Nuclear Fraction"

# Cell identity - UMAP + PCA
umapsm <- plotUMAP(sce_be1, colour_by = "Sample", dimred = "UMAP") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (PCA) colored by cell identity")
pcasm  <- plotPCA(sce_be1, colour_by = "Sample") +
  theme(legend.position = "none") +
  labs(title = "PCA colored by cell identity")

# Nuclear Fraction - UMAP + PCA
umapnf <- plotUMAP(sce_be1, colour_by = "nuclear_fraction", dimred = "UMAP") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (PCA) colored by nuclear fraction")
pcanf  <- plotPCA(sce_be1, colour_by = "nuclear_fraction") +
  theme(legend.position = "none") +
  labs(title = "PCA colored by nuclear fraction")
```

#### scGPT embeddings UMAP

```{r}
#| label: plot-umap-scgpt
umapgs <- plotUMAP(sce_be1, colour_by = "Sample", dimred = "UMAP_scGPT") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (scGPT) colored by cell identity")
umapgnf <- plotUMAP(sce_be1, colour_by = "nuclear_fraction", dimred = "UMAP_scGPT") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (scGPT) colored by nuclear fraction")
```

#### GeneFormer embeddings UMAP

```{r}
#| label: plot-umap-gf
umapgfs <- plotUMAP(sce_be1, colour_by = "Sample", dimred = "UMAP_GeneFormer") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (GeneFormer) colored by cell identity")
umapgfnf <- plotUMAP(sce_be1, colour_by = "nuclear_fraction", dimred = "UMAP_GeneFormer") +
  theme(legend.position = "bottom") +
  labs(title = "UMAP (GeneFormer) colored by nuclear fraction")
```

```{r}
#| label: plots-pca-results
#| fig-width: 12
#| fig-height: 10

(pcasm | pcanf ) / (umapsm | umapnf )
```

```{r}
#| label: plots-scgpt-gf
#| fig-width: 12
#| fig-height: 10
(umapgs | umapgnf) / (umapgfs | umapgfnf)
```

------------------------------------------------------------------------

## Focus on MALAT1 and Nuclear Fraction Relationship

We evaluate the behavior of the nuclear-retained long non-coding RNA **MALAT1** to biologically validate our nuclear fraction metric.

```{r}
#| label: malat1-analysis
#| fig-width: 12
#| fig-height: 5

umapm <- plotUMAP(sce_be1, colour_by = "MALAT1") + theme(legend.position = "bottom") +
  labs(title = "UMAP (PCA) colored by MALAT1 log expression")
umapgfm <- plotUMAP(sce_be1, colour_by = "MALAT1", dimred = "UMAP_GeneFormer") + theme(legend.position = "bottom") +
  labs(title = "UMAP (GeneFormer) colored by MALAT1")
umapgptm <- plotUMAP(sce_be1, colour_by = "MALAT1", dimred = "UMAP_scGPT") + theme(legend.position = "bottom") +
  labs(title = "UMAP (scGPT) colored by MALAT1")

# Direct comparison in reduced dimensions
umapm | umapgfm | umapgptm
```

### UMAP of scGPT embeddings

```{r}
#| label: comp-scgpt
#| fig-width: 12
#| fig-height: 5
umapgnf | umapgptm | umapgs
```

### UMAP of GeneFormer embeddings

```{r}
#| label: comp-gf
#| fig-width: 12
#| fig-height: 5
umapgfnf | umapgfm | umapgfs
```

```{r}
#| label: comp-malat1
#| fig-width: 12
#| fig-height: 5
umapm | umapnf | umapsm
```

### MALAT1 vs Nuclear Fraction Scatter Plot Faceted by Sample

```{r}
#| label: scatter-malat1-nf
#| fig-width: 12
#| fig-height: 6

plotExpression(sce_be1,
               features = "MALAT1",
               x = "nuclear_fraction",
               other_fields = "Sample",
               exprs_values = "logcounts") +
  facet_wrap(~Sample) +
  aes(color = Sample) +
  theme_minimal() +
  labs(title = "MALAT1 Expression vs Nuclear Fraction per Sample",
       x = "Nuclear Fraction",
       y = "MALAT1 Expression (logcounts)")
```

------------------------------------------------------------------------

### Detecting Empty Droplets and Damaged cells with `DropletQC`

We prepare the data frame for the `DropletQC` algorithm by handling missing values (`NA`) and evaluating the empty droplet cutoff.

```{r}
#| label: empty-droplets
#| fig-width: 8
#| fig-height: 6

nf_umi <- data.frame(
  nf = sce_be1$nuclear_fraction,
  umi = sce_be1$sum,
  Sample = sce_be1$Sample
)

rownames(nf_umi) <- colnames(sce_be1)

# Handle NA values
nf_umi <- nf_umi[!is.na(nf_umi$nf),]

# Identify empty droplets and output the diagnostic plot
cb_be1 <- identify_empty_drops(nf_umi = nf_umi, include_plot = TRUE)
```

```{r}
#| label: cell-status-prop
# Summary table of cell status
table(cb_be1$cell_status)
```

## Investigation of possible association between empty droplets, cellS identity and MALAT1 gene

Add the cell status column in the coldata of sce_be1

```{r}
#| label: cell_status-add
#| eval: false
sce_be1$cell_status <- cb_be1[colnames(sce_be1), "cell_status"]
```

```{r}
table(sce_be1$Sample, sce_be1$cell_status)
```

```{r}
p1 <- sce_be1[, !is.na(sce_be1$cell_status)] |>
  ggcells(aes(x = cell_status, y = nuclear_fraction, fill = cell_status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +  # Disegna il boxplot
  facet_wrap(~ Sample) +                           # Ora Sample è visibile!
  theme_minimal() +
  labs(
    title = "Nuclear fraction distribution for cell status",
    subtitle = "By each cell identity",
    x = "Cell Status",
    y = "Nuclear Fraction",
    fill = "Status"
  )

p2 <- sce_be1[, !is.na(sce_be1$cell_status)] |>
  ggcells(aes(x = cell_status, y = MALAT1, fill = cell_status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~ Sample) +
  theme_minimal() +
  labs(
    title = "MALAT1 distribution for cell status",
    subtitle = "By each cell identity",
    x = "Cell Status",
    y = "MALAT1 logexpression",
    fill = "Status"
  )

p1 | p2
```

### Damaged Cell Identification

Finally, we identify damaged cells by leveraging cell type annotations and save the finalized object.

```{r}
#| label: damaged-cells
#| fig-width: 10
#| fig-height: 12

# Reorder columns
cb_be1 <- cb_be1 |>
  relocate(Sample, .after = cell_status)

# Identify damaged cells using DropletQC
cb_be1_dc <- identify_damaged_cells(cb_be1, verbose = FALSE, output_plots = TRUE)

# Final cell status summary
table(cb_be1_dc[[1]]$cell_status)

# Diagnostic plots for each cell identity group
wrap_plots(cb_be1_dc[[2]], nrow = 6)
```

# Last plots: UMAP damaged and empty droplets

```{r}
#| label: umap-damaged-empty
#| fig-width: 10
#| fig-height: 11
umapc <- plotUMAP(sce_be1, colour_by = "cell_status") + theme(legend.position = "bottom") +
  labs(title = "UMAP (PCA) colored by cell status")
pcam  <- plotPCA(sce_be1, colour_by = "cell_status") + theme(legend.position = "bottom") +
  labs(title = "PCA colored by cell status")
umapec <- plotUMAP(sce_be1, colour_by = "cell_status", dimred = "UMAP_scGPT") + theme(legend.position = "bottom") +
  labs(title = "UMAP (scGPT) colored by cell status")
umapeg <- plotUMAP(sce_be1, colour_by = "cell_status", dimred = "UMAP_GeneFormer") + theme(legend.position = "bottom") +
  labs(title = "UMAP (GeneFormer) colored by cell status")

( umapc | pcam ) / ( umapec | umapeg )
```

and the relationship between cell status and MALAT1 expression in the embeddings

```{r}
#| label: scgpt-cs-malat1
#| fig-width: 12
#| fig-height: 5

umapec | umapgptm | umapgs
```

```{r}
#| label: gf-cs-malat1
#| fig-width: 12
#| fig-height: 5

umapeg | umapgfm | umapgfs
```

### Saving the Updated SCE Object

```{r}
#| label: save-object
#| eval: false


saveRDS(sce_be1, file = "/projects/shared/intronic_bam/datasets/BE1/BE1_sce.RDS")
```
